Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-04 main.Rmd

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates_ensembl_2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley_biomart_2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlov_bendtsen_improved_2004] in order to filter the secreted sequences and selected accordingly. ^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camacho_blast_2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

```{r effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt"}
knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf",
)
```

```{r noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists"}
knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
)
```

@@

^[Ruth, will we do a shiny tool?]

```{r effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model"}
knitr::include_graphics("figures/model_arch.png")
```

```{r trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set"}
knitr::include_graphics("figures/final_train_progress.png")
```

```{r overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
```

```{r confusionfinalmodels, eval=FALSE, include=FALSE, fig.cap="Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}

knitr::include_graphics("figures/final_confusion.png")
```

```{r modelcorrs, eval=FALSE, include=FALSE, fig.cap="Pearson correlations between classifications made by different models; A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
knitr::include_graphics("figures/drafts.png")
```

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
# Load data
bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")

@@

```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
```

```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

@@

```{r bacefftable}
apa_table(
  readr::read_tsv("tables/bacterial_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
)
```

```{r fungiefftable}
apa_table(
  readr::read_tsv("tables/fungal_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
)
```

```{r oomefftable}
apa_table(
  readr::read_tsv("tables/oomycete_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
)
```

```{r possiblecombination}
apa_table(
  readr::read_tsv("tables/possible_combination_each_model.tab"),
  digits = 0,
  align = 'c',
  caption = 'List of possible combination of hyperparamaters setting for each model'
)
```

```{r fixedhyperparam}
apa_table(
  readr::read_tsv("tables/fixed_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
)
```

```{r tunedhyperparam}
apa_table(
  readr::read_tsv("tables/tuned_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
)
```

```{r hyperparam}
 readr::read_tsv("tables/hyperparamscan_info.tab") %>% 
  knitr::kable( "latex", longtable = T, booktabs = T,
           digits = 0,
  align = 'c',
  caption = 'Hyperparameters used in automatic scanning of hyperparameter space and best performing parameters values for each model'
         ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)

>>>>

Sequence data were collected from the PHI-Base database version 4.8 [@phibase] by accessing a text dump of the data kindly prepared on request by the PHI-Base team, the file can be accessed at [https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv). The pipeline in Figure \@ref(fig:effectorcollection) outlines the steps used. We filtered plant effector proteins and their taxonomic groups and collected sequences from UniProt Release 2019_05, using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-data-new/binary-class/0001_first_step_getting_data.Rmd). We created a correspondingly sized data set of non-effectors with secretion signals originating in species matched to those from which the effectors were drawn. We downloaded sequences for randomly selected proteins matching these criteria from Ensembl databases [@yates2020] : specifically Ensembl Fungi, Protists and Bacteria manually using the BioMart tools [@smedley2009]. Since the BioMart tool is not available on Ensembl Bacteria, we downloaded whole proteome protein sequnces from species matched to those from which the effector came using FTP. With these we used SignalP 3.0 [@dyrlovbendtsen2004] in order to filter the secreted sequences and selected accordingly. ^[Ruth, what version of the code and what parameter settings did you use?] Redundant sequences were filtered using BLASTp [@camachoblast2009]. We achieved these steps using the code in [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/scripts/r-scripts/getting-secreted-data/0005_process_signalp_data.Rmd).

```{r, effectorcollection, out.width='50%', fig.align='center', fig.cap="Workflow diagram for collection of effector sequences from the PHI-Base database annotation and cross reference to UniProt", message = FALSE, echo=FALSE}

knitr::include_graphics(
  "figures/flowchart_effectorsequence_collection.pdf",
)
```

```{r, noneffectorcollection, fig.cap="Workflow diagram for collection of secreted non-effector sequences from Ensembl Bacteria, Fungi, and Protists", message = FALSE, echo=FALSE}

knitr::include_graphics(
  "figures/flowchart_secreted_noneffectorsequence_collection.pdf",
)
```

@@

```{r, effsummary, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r, modelarch, fig.cap="Overview of model architectures tested in this analysis, A: CNN-LSTM model; B: CNN-GRU model; C:LSTM-Embedding model; D:GRU-Embedding model", message = FALSE, echo=FALSE}
knitr::include_graphics("figures/model_arch.png")
```

```{r, trainingprogress, fig.cap="Training trajectories showing Accuracy and Loss over 40 epochs for four model types on A: Bacterial sequence training set; B: Fungal sequence training set; C: Oomycete sequence training set", message = FALSE, echo=FALSE}
knitr::include_graphics("figures/final_train_progress.png")
```

```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.", echo=FALSE, message=FALSE}

# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x") 
```

```{r confusionmodelcorrs, fig.height = 8, fig.width = 8, fig.cap = "Confusion matrices showing tendency of models to call false positive or false negative errors on the final test data; and Pearson correlations between classifications made by different models. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences.", message = FALSE, echo=FALSE}
# Load data
bacteria_ensemble_results <- data.table::fread("data/raw/bacteria_pred_ens.csv", drop = "V1")

@@

```{r ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.", message = FALSE, echo=FALSE}

# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")
```

```{r vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software", message = FALSE, echo=FALSE}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

@@

```{r, bacefftable, echo=FALSE, message = FALSE}

apa_table(
  readr::read_tsv("tables/bacterial_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Bacterial species contributing to the bacterial effector sequence set retrieved from PHI-Base'
)
```

```{r fungiefftable}

apa_table(
  readr::read_tsv("tables/fungal_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Fungal species contributing to the fungal effector sequence set retrieved from PHI-Base'
)
```

```{r, oomefftable, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/oomycete_effector_count.tab"),
  digits = 0,
  align = 'c',
  caption = 'Oomycete species contributing to the oomycete effector sequence set retrieved from PHI-Base'
)
```

```{r, possiblecombination, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/possible_combination_each_model.tab"),
  digits = 0,
  align = 'c',
  caption = 'List of possible combination of hyperparamaters setting for each model'
)
```

```{r, fixedhyperparam, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/fixed_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters fixed in automatic scanning of hyperparameter space for each model'
)
```

```{r, tunedhyperparam, echo=FALSE,message = FALSE}

apa_table(
  readr::read_tsv("tables/tuned_hyperparam.tab"),
  digits = 0,
  align = 'c',
  caption = 'Hyperparameters tuned in automatic scanning of hyperparameter space for each model'
)
```

```{r, hyperparam, echo=FALSE,message = FALSE }

 readr::read_tsv("tables/hyperparamscan_info.tab") %>% 
  knitr::kable( "latex", longtable = T, booktabs = T,
           digits = 0,
  align = 'c',
  caption = 'Hyperparameters used in automatic scanning of hyperparameter space and best performing parameters values for each model'
         ) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 7)
